One-Particle Density Matrix Polarisation Susceptibility Tensors

### 2018-08-13 papers/1_dmatpol/submit/paper-submit.tex
Citation added and switch to reprint mode

%%
%%   Version 3.1 of 31 July 2018.
%%
\documentclass[aip,amsmath,amssymb,reprint,floatfix]{revtex4-1}
%\documentclass[aip,reprint]{revtex4-1}

@@

\preprint{AIP/123-DMS}

@@

However, even with the modern advances in computational chemistry, accurate description of phenomena due 
to complex heterogeneous molecular environments remains a challenge in the following areas: (i) describing 
the electronically excited states (electronic solvatochromism),\cite{Barbati.JACS.2014,
Szabla.Sponer.Jiri.Gora.JPCL.2015,Bednarska.Zalesny.Tian.Murugan.Agren.Bartkowiak.Molecules.2017,
Jedrzejewska.Grabarz.Bartkowiak.Osmialowski.SpectChimActA.2018} 
(ii) understanding the vibrational properties of molecules in ground and electronically excited states
(vibrational or vibronic solvatochromisms).\cite{Blasiak.Londergan.Webb.Cho.ACR.2017,
Xu.Blasiak.Cho.Layfield.Londergan.JPCL.2018} 
Interface between QM and MM levels is usually realised by electrostatic embedding and only rarely 
the Pauli exclusion principle and dispersion are additionally included.\cite{List.Olsen.Kongsted.PCCP.2016}
Unfortunately, it is quite difficult to efficiently treat the environment\hyp{}induced potential operator
due to complicated nature of intermolecular interactions that theoretically cannot be solely envisioned 
as a sum of separate contributions coming from electrostatic, dispersion and Pauli 
effects.\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994} 
For example,
one of the most frequently used, efficient and reasonably accurate \emph{ab initio} fragmentation methods, 
the effective fragment potential (EFP) methods, are limited to the Hartree\hyp{}Fock 
(HF)\cite{Roothaan.RevModPhys.1951,Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013} 
or density functional theory (DFT) levels\cite{Hohenberg.Kohn.PhysRev.1964,Kohn.Sham.PhysRev.1965,Nguyen.Pachter.Day.JCP.2014}.
Unfortunately, while they can approximately describe the environment composed of molecules in their ground states,
they cannot be used for highly correlated wavefunctions such as electronically excited states.

@@

Although Eq.~\eqref{e:dmu-l-vector}
is exact up to first order in ${\BM\Delta}$, it cannot be inverted at this moment to obtain $\delta {\bf C}$.
If the orbitals are real, the analytic solution exists
provided that the electric field is uniform. In a more general case, i.e., when the electric field
is non\hyp{}uniform, the analytic solution cannot be found without further approximations.
Let the induced dipole moment be partitioned into separate contributions
from the occupied localised MO's (LMO's), i.e.,
%
\begin{equation} \label{e:dmu-part}
 \delta {\BM{\upmu}} = \sum_i^{\rm occ} \delta {\BM{\upmu}}_i \;,
\end{equation}
%
where the induced dipole moment associated with the $i$th LMO
is assumed to be linearly dependent on the local electric field, i.e.,
%
\begin{equation} \label{e:dmu-dpol}
 \delta {\BM{\upmu}}_i \approx {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) \;.
\end{equation}
%
In Eq.~\eqref{e:dmu-dpol}, ${\BM{\alpha}}_i$ is the distributed dipole\hyp{}dipole polarisability tensor
associated with the $i$th LMO whereas ${\bf F}({\bf r}_i)$ 
is the electric field evaluated at the $i$th LMO charge centroid,
${\bf r}_i = \tbraket{\phi_i}{\hat{\bf r}}{\phi_i}$.
By inserting the above result into Eq.~\eqref{e:dmu-l-vector},
recasting the right hand side in a matrix form
and assuming real orbitals,
one can separate the contributions coming from the occupied MO's as
%
\begin{equation} \label{e:dmu-l-vector-mo-transform-X}
 \frac{1}{4} {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) 
   =
   \delta {\bf c}_i^{\rm T} \cdot {\bf L}_i \;.
\end{equation}
%
Note that now ${\bf L}_i \equiv \left[ \mathbb{L} \right]_i$ 
is a real rectangular matrix of dimension $n \times 3$ and $\delta{\bf c}_i$ is the $i$th column of the
change in the LCAO-MO matrix due to the polarisation process.
The solution for $\delta {\bf C}$ is
%
\begin{equation} \label{e:mu-ind.matrix.linear.solution}
  \delta {\bf c}_i^{\rm T} = \frac{1}{4}
            {\bf F}({\bf r}_i)  \cdot {\BM{\alpha}}_i \cdot 
                    \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \;,
\end{equation}
%
where $\left[ {\bf L}_i  \right]^{-1}_{\rm Left}$ is a left inverse
of ${\bf L}_i$ matrix,
%
\begin{equation} 
      \left[ {\bf L}_i  \right]^{-1}_{\rm Left}   \equiv
       \left[ {\bf L}_i^{\rm T} \cdot {\bf L}_i \right]^{-1} \cdot {\bf L}_i^{\rm T}   
\end{equation}
%
with the dimension of $3\times n$. Note that the right inverse of ${\bf L}_i$
does not exist because ${\bf L}_i \cdot {\bf L}_i^{\rm T}$ is singular due to the fact that usually
$n > 3$.

@@

Now we can obtain the final expression for the density matrix change 
due to external electric field that reproduces the 
induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parameterised as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha}^{(i;1)} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha;w}^{(i;1)}= \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. 
The superscript `$(1)$' underlines the fact that the susceptibility introduced here
is first\hyp{}order with respect to the electric field.
The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i;1)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarisation susceptibility tensor} is
%
\begin{multline}  \label{e:susceptibility-B}
 {\bf B}^{(i;1)}_{\alpha\beta} = 
                               C_{\alpha i} {\bf b}_{\beta}^{(i;1)} + C_{\beta i} {\bf b}_\alpha^{(i;1)} \\
                                - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma} C_{\beta i} + D_{\beta\gamma} C_{\alpha i}
                                 \right)
                                           {\bf b}_{\gamma}^{(i;1)}
 \;.
\end{multline}
%
Hence, the three dimensional vector ${\bf B}^{(i;1)}_{\alpha\beta}$ 
describes the local linear polarisation affinity 
of the density matrix element 
associated with the $i$th LMO centroid as well as the $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasised here that all the elements of ${\bf B}_{\alpha\beta}^{(i;1)}$
are unique properties of electronically unperturbed molecule. 
However, they are valid only under the Hartree\hyp{}Fock
approximation, with a guarantee that only the overall charge (exactly) and the dipole moment (approximately,
limited by the flavour of the distributed polarisability approximation), 
but not higher multipole moments and polarisation energy, are correctly described.

@@

Eq.~\eqref{e:final-model.HF}, which is the final result of the previous section, 
can be used as a guideline to construct a general model of density matrix polarisation.
Quadratic (and higher\hyp{}order) terms with respect to the electric field
can be added to the formulation, which leads to
%
\begin{multline}\label{e:final-model.General}
 \delta D_{\alpha\beta} = \sum_{i }^M \Big\{
                                      {\bf B}^{(i ; 1)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \\
%                       + \sum_{ij}^M {\bf B}^{(ij; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_j)
                        +             {\bf B}^{(i ; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) 
                        + \ldots \Big\} \;.
\end{multline}
%
In the above equation, the density matrix susceptibilities
${\bf B}^{(i; R)}_{\alpha\beta}$, distributed over $M$ sites,
denote the $R$-th order response with respect to the electric field.
Note that
${\bf B}^{(i; 1)}_{\alpha\beta}$ is a vector of length $3$ whereas ${\bf B}^{(i; 2)}_{\alpha\beta}$
is a $3\times 3$ matrix. Also, the distributed expansion centres
(labelled as $i$) are generalised to any point in the molecule.\footnote{More general
contribution due to square of electric field could 
be $\sum_{ij}^M {\bf B}^{(ij; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_j)$. 
However, for simplicity, we neglect the contributions from different sites $i$ and $j$
in the present discussion and leave it for future development.
} Note that Eq.~\eqref{e:final-model.General} can refer to any type of one\hyp{}particle density matrix
such as ground\hyp{}state, excited\hyp{}state or transition density matrix.
However, the exact mathematical form of the generalised density matrix susceptibilities 
from Eq.~\eqref{e:final-model.General} is unknown.

@@

Let $\left\{ {\bf F}^{(1)}({\bf r}), {\bf F}^{(2)}({\bf r}), \ldots, {\bf F}^{(N)}({\bf r}), \ldots \right\}$ 
be a set of $N_{\rm max}$ distinct and randomly sampled 
spatial distributions of electric field. It is assumed that
the exact difference one\hyp{}particle density matrices (with respect to the unperturbed state)
defined as
%
\begin{equation}
 \delta \overline{\bf D}^{(N)} \equiv \overline{\bf D}^{(N)} - \overline{\bf D}^{(0)}
\end{equation}
%
are known for each sample (overline symbolises the exact estimate).
Now, for each pair of the AO indices we construct the following 
parameterisation of Eq.~\eqref{e:final-model.General}:
%
\begin{multline}\label{e:final-model.General.Parameters}
 \delta D^{(N)} = \sum_{i }^M \Big\{ 
                              \sum_u^{x,y,z} s^{[1]}_{iu} F_{iu}^{(N)} \\
%               + \sum_{ij}^M \sum_{uw}^{x,y,z} \left( s^{[2]}_{iu} + s^{[2]}_{jw} \right) F_{iu}^{(N)} F_{jw}^{(N)}
                +             \sum_{u}^{x,y,z} \sum_{w<u} r_{uw} s^{[2]}_{iuw} F_{iu}^{(N)} F_{iw}^{(N)} 
                        + \ldots \Big\}
\end{multline}
%
(the Greek subscripts were omitted here for notational simplicity).
In the above equation, $B_u^{(i;1)} = s^{[1]}_{iu}$ and $B_{uw}^{(i;2)} = r_{uw} s^{[2]}_{iuw}$,
where $r_{uw}$ is the symmetry factor equal to 1 for diagonal elements and 2 for off\hyp{}diagonal
elements of $B_{uw}^{(i;2)}$.
The multiple parameter blocks
(${\bf s}^{[1]}$, ${\bf s}^{[2]}$ and so on)
appear in the first power, allowing for linear least\hyp{}squares regression.
The square bracket superscripts denote the block of the parameter space.

@@

In order to validate all the models discussed in this contribution,
the electric field\hyp{}induced polarisation energy, 
dipole moment and quadrupole moment were analysed in detail.
The electric field\hyp{}induced polarisation energy at the HF level, $\delta E^{\rm Pol}$,
was computed from the following exact formula:
%
\begin{multline}\label{e:dE}
 \delta E^{\rm Pol}[{\bf F}] = 
                     \Tr{\left\{ {\bf D}^{(0)} \cdot
                                \left( 2\delta {\bf J} - \delta {\bf K}  \right) \right\}} \\
                   + \Tr{\left\{ \delta {\bf D} \cdot
                                \left( {\bf H}^{(0)} + {\bf f}^{(0)} + 2{\bf V} + 2\delta {\bf J} - \delta {\bf K}\right) \right\}} \;.
\end{multline}
%
In Eq.~\eqref{e:dE}, ${\bf f}^{(0)}$ and ${\bf H}^{(0)}$ are the Fock matrix and 
the one\hyp{}electron core Hamiltonian matrix of an unperturbed HF ground state,
${\bf V}$ is the one\hyp{}electron external electrostatic potential matrix whereas $\delta {\bf J}$
and $\delta {\bf K}$ are electric field\hyp{}induced changes in the Coulomb and the exchange HF matrices due to 
the polarisation of the electron density, respectively. 
In order to compare the performance of the usual 
distributed dipole\hyp{}dipole polarisability approach,\cite{Li.Netzloff.Gordon.JCP.2006}
the approximate interaction energy was also evaluated by using the expression:
%
\begin{equation}\label{e:dE-dpol}
 \delta E^{\rm Pol}[{\bf F}] \approx 
                   - \frac{1}{2} \sum_i^{\rm occ} {\BM \alpha}_i : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) \;,
\end{equation}
%
in which ${\BM \alpha}_i$ is the distributed anisotropic dipole\hyp{}dipole polarisability tensor associated with the
$i$th LMO centroid (see Eq.~\eqref{e:dmu-dpol}).
The distributed polarisabilities were computed by utilising the coupled\hyp{}perturbed Hartree\hyp{}Fock
method.\cite{McWeeny.RevModPhys.1960,Dodds.McWeeney.Sadlej.MolPhys.1977} 
Orbital localisation was performed by using the Boys method.\cite{Boys.RevModPhys.1960}

@@

Polarisation under the uniform electric field is the simplest type of electrostatic perturbation
whereby it is possible to compute the exact DMS according to Eq.~\eqref{e:dD-Taylor} 
directly by numerical differentiation. Here we consider DMS up to second order, i.e.,
%
\begin{subequations}\label{e:exact-DMS}
 \begin{align}
  {\bf B}^{(10)}_{\alpha\beta} &= \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B10} \\
  {\bf B}^{(20)}_{\alpha\beta} &= \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B20}
 \end{align}
\end{subequations}
%
and electric field magnitudes in a range from very weak ($<$0.001 a.u.) to very strong ($\sim$0.07 a.u.).
The \emph{ab initio} model from Eq.~\eqref{e:final-model.HF} and
the generalised model from Eq.~\eqref{e:final-model.General} are also discussed for the sake of comparison. 
The associated error distributions in prediction of polarisation energies 
are shown in scatter plot in Figure~\ref{f:fig-1}, whereas the
statistical performance descriptors of polarisation energies and induced multipole moments
are gathered in Table~\ref{t:results}.
%
\begin{figure}[t]
\includegraphics[width=0.5\textwidth]{fig-1.eps}
\caption{\label{f:fig-1} {\bf Performance of the DMS tensors of the ground\hyp{}state water molecule
in uniform electric field.} 
$R_{\rm max}$ denotes the maximum rank of the electric field
in the expansion from Eq.~\eqref{e:final-model.General}.
Electric field magnitudes in a range from 0.0 to 0.07 a.u. were tested in 100 samples with randomly selected 
directions of electric field.
} 
\end{figure}
%

>>>>

%%
%%   Version 3.1 of 31 July 2018.
%%
\documentclass[aip,jcp,preprint,amsmath,amssymb,floatfix]{revtex4-1}
%\documentclass[aip,reprint]{revtex4-1}

@@

%\preprint{AIP/123-DMS}

@@

However, even with the modern advances in computational chemistry, accurate description of phenomena due 
to complex heterogeneous molecular environments remains a challenge in the following areas: (i) describing 
the electronically excited states (electronic solvatochromism),\cite{Barbati.JACS.2014,
Szabla.Sponer.Jiri.Gora.JPCL.2015,Bednarska.Zalesny.Tian.Murugan.Agren.Bartkowiak.Molecules.2017,
Jedrzejewska.Grabarz.Bartkowiak.Osmialowski.SpectChimActA.2018} 
(ii) understanding the vibrational properties of molecules in ground and electronically excited states
(vibrational or vibronic solvatochromisms).\cite{Blasiak.Londergan.Webb.Cho.ACR.2017,
Xu.Blasiak.Cho.Layfield.Londergan.JPCL.2018,Lewis.Gruenke.Oliver.Ballottari.Bassi.Fleming.JPCL.2016} 
Interface between QM and MM levels is usually realised by electrostatic embedding and only rarely 
the Pauli exclusion principle and dispersion are additionally included.\cite{List.Olsen.Kongsted.PCCP.2016}
Unfortunately, it is quite difficult to efficiently treat the environment\hyp{}induced potential operator
due to complicated nature of intermolecular interactions that theoretically cannot be solely envisioned 
as a sum of separate contributions coming from electrostatic, dispersion and Pauli 
effects.\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994} 
For example,
one of the most frequently used, efficient and reasonably accurate \emph{ab initio} fragmentation methods, 
the effective fragment potential (EFP) methods, are limited to the Hartree\hyp{}Fock 
(HF)\cite{Roothaan.RevModPhys.1951,Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013} 
or density functional theory (DFT) levels\cite{Hohenberg.Kohn.PhysRev.1964,Kohn.Sham.PhysRev.1965,Nguyen.Pachter.Day.JCP.2014}.
Unfortunately, while they can approximately describe the environment composed of molecules in their ground states,
they cannot be used for highly correlated wavefunctions such as electronically excited states.

@@

Although Eq.~\eqref{e:dmu-l-vector}
is exact up to first order in ${\BM\Delta}$, it cannot be inverted at this moment to obtain $\delta {\bf C}$.
If the orbitals are real, the analytic solution exists
provided that the electric field is uniform. In a more general case, i.e., when the electric field
is non\hyp{}uniform, the analytic solution cannot be found without further approximations.
Let the induced dipole moment be partitioned into separate contributions
from the occupied localised MO's (LMO's), i.e.,
%
\begin{equation} \label{e:dmu-part}
 \delta {\BM{\upmu}} = \sum_i^{\rm occ} \delta {\BM{\upmu}}_i \;,
\end{equation}
%
where the induced dipole moment associated with the $i$th LMO
is assumed to be linearly dependent on the local electric field, i.e.,
%
\begin{equation} \label{e:dmu-dpol}
 \delta {\BM{\upmu}}_i \approx {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) \;.
\end{equation}
%
In Eq.~\eqref{e:dmu-dpol}, ${\BM{\alpha}}_i$ is the distributed dipole\hyp{}dipole polarisability tensor
associated with the $i$th LMO whereas ${\bf F}({\bf r}_i)$ 
is the electric field evaluated at the $i$th LMO charge centroid,
${\bf r}_i = \tbraket{\phi_i}{\hat{\bf r}}{\phi_i}$.
By inserting the above result into Eq.~\eqref{e:dmu-l-vector},
recasting the right hand side in a matrix form
and assuming real orbitals,
one can separate the contributions coming from the occupied MO's as
%
\begin{equation} \label{e:dmu-l-vector-mo-transform-X}
 \frac{1}{4} {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) 
   =
   \delta {\bf c}_i^{\rm T} \cdot {\bf L}_i \;.
\end{equation}
%
Note that now ${\bf L}_i \equiv \left[ \mathbb{L} \right]_i$ 
is a real rectangular matrix of dimension $n \times 3$ and $\delta{\bf c}_i$ is the $i$th column of the
change in the LCAO-MO matrix due to the polarisation process.
The solution for $\delta {\bf C}$ is
%
\begin{equation} \label{e:mu-ind.matrix.linear.solution}
  \delta {\bf c}_i^{\rm T} = \frac{1}{4}
                    {\bf F}^{\rm T}({\bf r}_i) \cdot {\BM{\alpha}}_i^{\rm T} \cdot
                    \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \;,
%                   {\BM{\alpha}}_i : {\bf F}({\bf r}_i)  \otimes
%                   \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \;,
\end{equation}
%
where $\left[ {\bf L}_i  \right]^{-1}_{\rm Left}$ is a left inverse
of ${\bf L}_i$ matrix,
%
\begin{equation} 
      \left[ {\bf L}_i  \right]^{-1}_{\rm Left}   \equiv
       \left[ {\bf L}_i^{\rm T} \cdot {\bf L}_i \right]^{-1} \cdot {\bf L}_i^{\rm T}   
\end{equation}
%
with the dimension of $3\times n$. Note that the right inverse of ${\bf L}_i$
does not exist because ${\bf L}_i \cdot {\bf L}_i^{\rm T}$ is singular due to the fact that usually
$n > 3$.

@@

Now we can obtain the final expression for the density matrix change 
due to external electric field that reproduces the 
induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parameterised as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha}^{(i;1)} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha;w}^{(i;1)}= \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. 
The superscript `$(1)$' underlines the fact that the susceptibility introduced here
is first\hyp{}order with respect to the electric field.
The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i;1)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarisation susceptibility tensor} is
%
%\begin{multline}  \label{e:susceptibility-B} % PREPRINT
\begin{equation}  \label{e:susceptibility-B}
 {\bf B}^{(i;1)}_{\alpha\beta} = 
                               C_{\alpha i} {\bf b}_{\beta}^{(i;1)} + C_{\beta i} {\bf b}_\alpha^{(i;1)} %\\ PREPRINT
                                - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma} C_{\beta i} + D_{\beta\gamma} C_{\alpha i}
                                 \right)
                                           {\bf b}_{\gamma}^{(i;1)}
 \;.
\end{equation}
%\end{multline} % PREPRINT
%
Hence, the three dimensional vector ${\bf B}^{(i;1)}_{\alpha\beta}$ 
describes the local linear polarisation affinity 
of the density matrix element 
associated with the $i$th LMO centroid as well as the $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasised here that all the elements of ${\bf B}_{\alpha\beta}^{(i;1)}$
are unique properties of electronically unperturbed molecule. 
However, they are valid only under the Hartree\hyp{}Fock
approximation, with a guarantee that only the overall charge (exactly) and the dipole moment (approximately,
limited by the flavour of the distributed polarisability approximation), 
but not higher multipole moments and polarisation energy, are correctly described.

@@

Eq.~\eqref{e:final-model.HF}, which is the final result of the previous section, 
can be used as a guideline to construct a general model of density matrix polarisation.
Quadratic (and higher\hyp{}order) terms with respect to the electric field
can be added to the formulation, which leads to
%
%\begin{multline}\label{e:final-model.General} % PREPRINT
\begin{equation}\label{e:final-model.General}
 \delta D_{\alpha\beta} = \sum_{i }^M \Big\{
                                      {\bf B}^{(i ; 1)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  %\\ PREPRINT
%                       + \sum_{ij}^M {\bf B}^{(ij; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_j)
                        +             {\bf B}^{(i ; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) 
                        + \ldots \Big\} \;.
\end{equation}
%\end{multline} % PREPRINT
%
In the above equation, the density matrix susceptibilities
${\bf B}^{(i; R)}_{\alpha\beta}$, distributed over $M$ sites,
denote the $R$-th order response with respect to the electric field.
Note that
${\bf B}^{(i; 1)}_{\alpha\beta}$ is a vector of length $3$ whereas ${\bf B}^{(i; 2)}_{\alpha\beta}$
is a $3\times 3$ matrix. Also, the distributed expansion centres
(labelled as $i$) are generalised to any point in the molecule.\footnote{More general
contribution due to square of electric field could 
be $\sum_{ij}^M {\bf B}^{(ij; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_j)$. 
However, for simplicity, we neglect the contributions from different sites $i$ and $j$
in the present discussion and leave it for future development.
} Note that Eq.~\eqref{e:final-model.General} can refer to any type of one\hyp{}particle density matrix
such as ground\hyp{}state, excited\hyp{}state or transition density matrix.
However, the exact mathematical form of the generalised density matrix susceptibilities 
from Eq.~\eqref{e:final-model.General} is unknown.

@@

Let $\left\{ {\bf F}^{(1)}({\bf r}), {\bf F}^{(2)}({\bf r}), \ldots, {\bf F}^{(N)}({\bf r}), \ldots \right\}$ 
be a set of $N_{\rm max}$ distinct and randomly sampled 
spatial distributions of electric field. It is assumed that
the exact difference one\hyp{}particle density matrices (with respect to the unperturbed state)
defined as
%
\begin{equation}
 \delta \overline{\bf D}^{(N)} \equiv \overline{\bf D}^{(N)} - \overline{\bf D}^{(0)}
\end{equation}
%
are known for each sample (overline symbolises the exact estimate).
Now, for each pair of the AO indices we construct the following 
parameterisation of Eq.~\eqref{e:final-model.General}:
%
%\begin{multline}\label{e:final-model.General.Parameters} % PREPRINT
\begin{equation}\label{e:final-model.General.Parameters}
 \delta D^{(N)} = \sum_{i }^M \Big\{ 
                              \sum_u^{x,y,z} s^{[1]}_{iu} F_{iu}^{(N)} %\\ % PREPRINT
%               + \sum_{ij}^M \sum_{uw}^{x,y,z} \left( s^{[2]}_{iu} + s^{[2]}_{jw} \right) F_{iu}^{(N)} F_{jw}^{(N)}
                +             \sum_{u}^{x,y,z} \sum_{w<u} r_{uw} s^{[2]}_{iuw} F_{iu}^{(N)} F_{iw}^{(N)} 
                        + \ldots \Big\}
\end{equation}
%\end{multline} % PREPRINT
%
(the Greek subscripts were omitted here for notational simplicity).
In the above equation, $B_u^{(i;1)} = s^{[1]}_{iu}$ and $B_{uw}^{(i;2)} = r_{uw} s^{[2]}_{iuw}$,
where $r_{uw}$ is the symmetry factor equal to 1 for diagonal elements and 2 for off\hyp{}diagonal
elements of $B_{uw}^{(i;2)}$.
The multiple parameter blocks
(${\bf s}^{[1]}$, ${\bf s}^{[2]}$ and so on)
appear in the first power, allowing for linear least\hyp{}squares regression.
The square bracket superscripts denote the block of the parameter space.

@@

In order to validate all the models discussed in this contribution,
the electric field\hyp{}induced polarisation energy, 
dipole moment and quadrupole moment were analysed in detail.
The electric field\hyp{}induced polarisation energy at the HF level, $\delta E^{\rm Pol}$,
was computed from the following exact formula:
%
%\begin{multline}\label{e:dE} % PREPRINT
\begin{equation}\label{e:dE}
 \delta E^{\rm Pol}[{\bf F}] = 
                     \Tr{\left\{ {\bf D}^{(0)} \cdot
                                \left( 2\delta {\bf J} - \delta {\bf K}  \right) \right\}} %\\ % PREPRINT
                   + \Tr{\left\{ \delta {\bf D} \cdot
                                \left( {\bf H}^{(0)} + {\bf f}^{(0)} + 2{\bf V} + 2\delta {\bf J} - \delta {\bf K}\right) \right\}} \;.
\end{equation}
%\end{multline} % PREPRINT
%
In Eq.~\eqref{e:dE}, ${\bf f}^{(0)}$ and ${\bf H}^{(0)}$ are the Fock matrix and 
the one\hyp{}electron core Hamiltonian matrix of an unperturbed HF ground state,
${\bf V}$ is the one\hyp{}electron external electrostatic potential matrix whereas $\delta {\bf J}$
and $\delta {\bf K}$ are electric field\hyp{}induced changes in the Coulomb and the exchange HF matrices due to 
the polarisation of the electron density, respectively. 
In order to compare the performance of the usual 
distributed dipole\hyp{}dipole polarisability approach,\cite{Li.Netzloff.Gordon.JCP.2006}
the approximate interaction energy was also evaluated by using the expression:
%
\begin{equation}\label{e:dE-dpol}
 \delta E^{\rm Pol}[{\bf F}] \approx 
                   - \frac{1}{2} \sum_i^{\rm occ} {\BM \alpha}_i : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) \;,
\end{equation}
%
in which ${\BM \alpha}_i$ is the distributed anisotropic dipole\hyp{}dipole polarisability tensor associated with the
$i$th LMO centroid (see Eq.~\eqref{e:dmu-dpol}).
The distributed polarisabilities were computed by utilising the coupled\hyp{}perturbed Hartree\hyp{}Fock
method.\cite{McWeeny.RevModPhys.1960,Dodds.McWeeney.Sadlej.MolPhys.1977} 
Orbital localisation was performed by using the Boys method.\cite{Boys.RevModPhys.1960}

@@

Polarisation under the uniform electric field is the simplest type of electrostatic perturbation
whereby it is possible to compute the exact DMS according to Eq.~\eqref{e:dD-Taylor} 
directly by numerical differentiation. Here we consider DMS up to second order, i.e.,
%
\begin{subequations}\label{e:exact-DMS}
 \begin{align}
  {\bf B}^{(1)}_{\alpha\beta} &= \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B10} \\
  {\bf B}^{(2)}_{\alpha\beta} &= \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B20}
 \end{align}
\end{subequations}
%
and electric field magnitudes in a range from very weak ($<$0.001 a.u.) to very strong ($\sim$0.07 a.u.).
The \emph{ab initio} model from Eq.~\eqref{e:final-model.HF} and
the generalised model from Eq.~\eqref{e:final-model.General} are also discussed for the sake of comparison. 
The associated error distributions in prediction of polarisation energies 
are shown in scatter plot in Figure~\ref{f:fig-1}, whereas the
statistical performance descriptors of polarisation energies and induced multipole moments
are gathered in Table~\ref{t:results}.
%
\begin{figure}[t]
\includegraphics[width=0.5\textwidth]{fig-1.eps}
\caption{\label{f:fig-1} {\bf Performance of the DMS tensors of the ground\hyp{}state water molecule
in uniform electric field.} 
$R_{\rm max}$ denotes the maximum rank of the electric field
in the expansion from Eq.~\eqref{e:final-model.General}.
Electric field magnitudes in a range from 0.0 to 0.07 a.u. were tested in 100 samples with randomly selected 
directions of electric field.
} 
\end{figure}
%

All the tested models that are linear with respect to the electric field,
i.e., from Eq.~\eqref{e:final-model.HF} (black solid squares in Figure~\ref{f:fig-1}) 
as well as truncated at linear susceptibilities $\mathbb{B}^{(1)}$ 
from Eqs.~\eqref{e:dD-Taylor} (blue solid circles) and \eqref{e:final-model.General} (blue open circles),
incorrectly reproduce the polarisation energy.
Interestingly, the predicted values are in strong
correlation with the exact estimates and exhibit similar slope of $\sim$2.
Incorporation of the second\hyp{}order term
obtained either by numerical differentiation from Eq.~\eqref{e:B20} (red filled circles) 
or by using the generalised DMS model (red open circles)
dramatically reduces the errors
leading to quantitative accuracy in all tested field magnitudes 
with the best RMSE of 0.173 kcal/mol ($R^2=99.1\%$) for the generalised DMS model.
On the other hand, the
induced dipole and quadrupole moments are reproduced reasonably well by
using the first\hyp{}order DMS with $R^2$ coefficients being 
greater than 97\% in case of dipole moments and from 69\% up to 81\% in case of quadrupole moments.
This observation can be understood given the fact that the induced dipole
and quadrupole moment is approximately proportional to the external electric field, whereas the polarisation
energy exhibits quadratic dependence (c.f. Eq.~\eqref{e:dE-dpol}), hence the need of including at least
quadratic terms in the expansion from Eq.~\eqref{e:dD-Taylor} or Eq.~\eqref{e:final-model.General}.
